
# ============================================================
# An Analysis of Rural-Urban Differences in
# Drinking Water System Violations in Georgia
# Author: Talia C. Swanson
# ============================================================

#Project Part 1

Here we load all necessary R packages for data wrangling, plotting, and tidying model results.
# -----------------------------
# Load Required Packages
# -----------------------------
library(tidyverse)
library(here)
library(broom)
library(ggplot2)

# 1. LOAD DATA
# ----------------------------------------------------------------------------
This will import the public water system data, violation records, and rural-urban ZIP code classification for Georgia.
```{r}
systems <- read_csv(here("data/raw-data/SDWA_PUB_WATER_SYSTEMS.csv"))
violations <- read_csv(here("data/raw-data/SDWA_PN_VIOLATION_ASSOC.csv"))
ruca <- read_csv(here("data/raw-data/RUCA-codes-2020-zipcode.csv"))
```

# 2. FILTER FOR GEORGIA SYSTEMS
# -----------------------------------------------------------------------------

This will subset the data to only include water systems located in Georgia and check the resulting sample size.
```{r}
systems_ga <- systems %>%
  filter(STATE_CODE == "GA")

nrow(systems_ga)   # sample size check
```

# 3. EXPLORATORY DATA ANALYSIS
# -----------------------------------------------------------------------------
This step summarizes the types of water systems present and visualize their counts with a bar plot 
```{r}
# ---- System Type Counts ----
systems_ga %>%
  count(PWS_TYPE_CODE, sort = TRUE)

ggplot(systems_ga, aes(x = PWS_TYPE_CODE)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Count of Public Water System Types in Georgia",
    x = "Water System Type",
    y = "Number of Systems"
  )
```


# 4. POPULATION SERVED SUMMARY
This will examine the distribution of the population served across water systems. A log-scaled histogram shows that most systems serve relatively small populations, with a few large outliers

```{r}
# ---- Population Summary ----
summary(systems_ga$POPULATION_SERVED_COUNT)

# ---- Population Distribution ----
systems_ga_filtered <- systems_ga %>%
  filter(POPULATION_SERVED_COUNT > 0)

ggplot(systems_ga_filtered,
       aes(x = POPULATION_SERVED_COUNT)) +
  geom_histogram(bins = 40) +
  scale_x_log10(
    breaks = c(10, 100, 1000, 10000, 100000),
    labels = c("10", "100", "1,000", "10,000", "100,000")
  ) +
  labs(
    title = "Distribution of Population Served by Public Water Systems in Georgia",
    x = "Population Served",
    y = "Number of Systems"
  )
```

# 5. CREATE SYSTEM-LEVEL VIOLATION COUNTS
# -----------------------------------------------------------------------------
Here we examine the distribution of the population served across water systems. A log-scaled histogram shows that most systems serve relatively small populations, with a few large outliers
```{r}
violations_ga <- violations %>%
  inner_join(systems_ga %>% select(PWSID), by = "PWSID")

violation_counts <- violations_ga %>%
  group_by(PWSID) %>%
  summarise(violation_count = n(), .groups = "drop")

violation_summary <- violation_counts %>%
  count(violation_count) %>%
  rename(Number_of_Systems = n,
         Violation_Count = violation_count)

print(violation_summary)

ggplot(violation_summary,
       aes(x = factor(Violation_Count),
           y = Number_of_Systems)) +
  geom_bar(stat = "identity", fill = "darkred") +
  labs(
    title = "Distribution of Drinking Water Violations per Public Water System in Georgia",
    x = "Number of Violations",
    y = "Number of Water Systems"
  ) +
  theme_minimal()
```


#Project Part 2

# 6. MERGE VIOLATIONS BACK TO SYSTEM DATA
# -----------------------------------------------------------------------------
Violation records are aggregated to the system level. We summarize and plot how many systems have zero, one, or multiple violations.

```{r}
analysis_data <- systems_ga %>%
  left_join(violation_counts, by = "PWSID") %>%
  mutate(
    violation_count = ifelse(is.na(violation_count), 0, violation_count)
  )
```


# 7. MERGE RUCA & RURAL-URBAN CLASSIFICATION
# -----------------------------------------------------------------------------
We join RUCA ZIP code classifications to label each system as rural or urban. Missing values are removed.
```{r}
analysis_data <- analysis_data %>%
  mutate(ZIP_CODE = as.character(ZIP_CODE))

ruca <- ruca %>%
  mutate(ZIPCode = as.character(ZIPCode))

analysis_data <- analysis_data %>%
  left_join(ruca, by = c("ZIP_CODE" = "ZIPCode")) %>%
  mutate(
    rural = ifelse(PrimaryRUCA >= 4, 1, 0)
  )

table(analysis_data$rural, useNA = "ifany")

analysis_data <- analysis_data %>%
  filter(!is.na(rural))
```

# 8. CREATE BINARY VIOLATION OUTCOME
# -----------------------------------------------------------------------------
We join RUCA ZIP code classifications to label each system as rural or urban. Missing values are removed.
```{r}
analysis_data <- analysis_data %>%
  mutate(
    any_violation = ifelse(violation_count > 0, 1, 0)
  )

table(analysis_data$any_violation)
prop.table(table(analysis_data$any_violation))
```

# 9. PREPARE VARIABLES
# -----------------------------------------------------------------------------
A binary variable is created to indicate whether a system has any violations. We also check the distribution of this variable.
```{r}
analysis_data <- analysis_data %>%
  filter(POPULATION_SERVED_COUNT > 0) %>%
  mutate(
    log_population = log(POPULATION_SERVED_COUNT),
    rural = as.factor(rural),
    any_violation = as.factor(any_violation),
    PWS_TYPE_CODE = as.factor(PWS_TYPE_CODE)
  )

summary(analysis_data$log_population)

# Save processed dataset
write_csv(
  analysis_data,
  here("data/processed-data/analysis_data.csv")
)
```

# 10. FIT LOGISTIC REGRESSION MODEL
# ------------------------------------------------ 
Variables are transformed for analysis. Population is log-transformed, and categorical variables are set for modeling.
```{r}
model1 <- glm(
  any_violation ~ rural + log_population + PWS_TYPE_CODE,
  data = analysis_data,
  family = binomial
)

summary(model1)
```

# 10. ODDS RATIOS + 95% CI
# -----------------------------------------------------------------------------
The model output is converted to odds ratios with confidence intervals and saved for reporting.
```{r}
model_results <- tidy(
  model1,
  exponentiate = TRUE,
  conf.int = TRUE
)

print(model_results)

write_csv(
  model_results,
  here("data/processed-data/logistic_regression_results.csv")
)
```

# 11. FOREST PLOT
# -----------------------------------------------------------------------------
The model output is converted to odds ratios with confidence intervals and saved for reporting.
```{r}
model_plot_data <- model_results %>%
  filter(term != "(Intercept)")

forest_plot <- ggplot(model_plot_data,
                      aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low,
                    ymax = conf.high),
                width = 0.2) +
  coord_flip() +
  labs(
    title = "Odds Ratios for Logistic Regression Predictors",
    x = "Predictor",
    y = "Odds Ratio (95% CI)"
  ) +
  theme_minimal()

print(forest_plot)

ggsave(
  filename = here("products/manuscript/forest_plot.png"),
  plot = forest_plot,
  width = 8,
  height = 6
)
```


