
# ============================================================
# An Analysis of Rural-Urban Differences in
# Drinking Water System Violations in Georgia
# Author: Talia C. Swanson
# ============================================================

This section performs exploratory data analysis to better understand the structure o f the SDWIS public water systems dataset and examine basic characteristics of water systems in Georgia.

## Load Packages + Data
```{r}
library(tidyverse)
library(here)
library(knitr)

# Load datasets
systems <- read_csv(here("data/raw-data/GA_PUB_WATER_SYSTEMS.csv"))
ruca <- read_csv(here("data/raw-data/RUCA-codes-2020-zipcode.csv"))
```

The structure of the datasets is inspected to understand the variables available for analysis.

```{r}
glimpse(systems)
glimpse(ruca)
```

## Confirm Georgia water systems

```{r}
systems_ga <- systems

# Check sample size
nrow(systems_ga)
```

## Explore water system types

Next, the distribution of different types of public water systems in Georgia is examined.

``` {r}
ggplot(systems_ga, aes(x= PWS_TYPE_CODE))+
  geom_bar (fill="steelblue")+
  labs(
    title = "Count of Public Water System Types in Georgia",
    x = "Water System Type",
    y = "Number of Systems"
  )
```

```{r}
systems_ga |>
  count(PWS_TYPE_CODE, sort = TRUE)
```

## Population served summary and distribution

Summary statistics are used to understand the distribution and range of populations served by water systems.
summary

```{r}(systems_ga$POPULATION_SERVED_COUNT)
```

The distribution of population served is visualized using a histogram.

```{r}
systems_ga_filtered <- systems_ga |>
  filter(POPULATION_SERVED_COUNT>0)

ggplot(
  data=systems_ga_filtered,
  mapping = aes(x=POPULATION_SERVED_COUNT)
) +
  geom_histogram(bins=40)+
  scale_x_log10(
    breaks = c(10,100,1000,10000,100000),
    labels = c("10", "100", "1,000", "10,000", "100,000")
  ) +
  labs(
    title = "Distribution of Population Served by Public Water Systems in Georgia",
    x= "Population Served",
    y= "Number of Systems"
  )
```

## Violation dataset exploration

```{r}
systems_ga_filtered <- systems_ga |>
  filter(POPULATION_SERVED_COUNT>0)

ggplot(
  data=systems_ga_filtered,
  mapping = aes(x=POPULATION_SERVED_COUNT)
) +
  geom_histogram(bins=40)+
  scale_x_log10(
    breaks = c(10,100,1000,10000,100000),
    labels = c("10", "100", "1,000", "10,000", "100,000")
  ) +
  labs(
    title = "Distribution of Population Served by Public Water Systems in Georgia",
    x= "Population Served",
    y= "Number of Systems"
  )
```

## Count violations

The number of violations per system is summarized to understand how frequently violations occur across water systems.

```{r}
violation_summary <- violations_ga %>%
  group_by(PWSID) %>%
  summarise(violation_count = n()) %>%
  count(violation_count) %>%
  rename(Number_of_Systems = n, Violation_Count = violation_count)

print(violation_summary)
```

## Violation distribution plot

The distribution of violations across water systems is visualized using a bar plot.

```{r}
ggplot(violation_summary, aes(x = factor(Violation_Count), y = Number_of_Systems)) +
  geom_bar(stat = "identity", fill = "darkred") +
  labs(
    title = "Distribution of Drinking Water Violations per Public Water System in Georgia",
    x = "Number of Violations",
    y = "Number of Water Systems"
  ) +
  theme_minimal()
```
