---
title: "An Analysis of Rural-Urban Difference in Drinking Water System Violations in Georgia"
author: "Talia C. Swanson, B.S."
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

## Description of data and data source
The primary data source for this project is the Safe Drinking Water Infromation System (SDWIS) maintianed by the Environmental Protection Agency (EPA). SDWIS is a national databased maintained by the EPA to monitor compliance with the Safe Drinking Water Act. The data are collected through reports submitted by public water systems and state primacy agencies, including detailed information on water system characteristics and drinking water violations.

Two primary SDWIS datasets will be used with the first containg water system-level information inclduing system identification, numbers, system type, population served, geographic location, and ownership characteristics. The second dataset includes violation level records, documenting containment, violation type and dates.

## Questions/Hypotheses to be addressed
The primary research question for this project is:

Are rural public water systems in Georgia more likely to experience drinking water quality violations compared to urban water systems?

The outcome of interest is the occurence of drinking water quality violations (either as a binary indicator or collective count of violations per system). The primary predictor of interest is rural versus urban area classification of the water system's service area. The analysis will explore patterns in violation frequency and distribution across rural and urban systems and assess whether disparities persist after accounting for system and community characteristics. 



{{< pagebreak >}}


# Methods/Planned Analysis Approach ( Part 1 )

The analysis will begin with data cleaning and processing, including filtering for Georgia water systems, merging system and violation datasets, and handling missing values. Exploratory data analysis will be used to examine distributions, identify patterns, and visualize difference in violations between rural and urban systems. 

Descriptive summaries will be used to examine the distribution of water system characteristics such as population served and rural urban classfication. Visualization will be used to compare drinking water violations between rural and urban systems, such technique will include bar plots showing the proportions of systems with violations and histograms of violation counts and population service. 


# Sample Exploratory Data Analysis 
This will load the SDWA public water systems datatset and the RUCA ZIP code classification datatser into R. These datasets contian infromation on public water system characterstics and rural-urban classifications. 
```{r}

library(tidyverse)
library(here)


# Load datasets
systems <- read_csv(here("data/raw-data/SDWA_PUB_WATER_SYSTEMS.csv"))
ruca <- read_csv(here("data/raw-data/RUCA-codes-2020-zipcode.csv"))

glimpse(systems)
glimpse(ruca)

```


This will filter the public water system dataset to include only systems located in GA using the STATE_CODE variable. After filtering, there are 5256 remaining observations, confirming that the restriction was applied and to gain the final sample size for analysis. 
```{r}
systems_ga <- systems |>
  filter(STATE_CODE == "GA")

# Check sample size
nrow(systems_ga)
```

```{r}
systems_ga_filtered <- systems_ga
```

Looking at the types of public water systems in Georgia, community water systems were the most common (n=2722), followed by transient non-community systems n= 2,004 and non-transiet non-community systems (n=530)
  * CWS (community water system)
  * TNCWS (transient non-community water system)
  * NTNCWS (non-transient non-community water system)

```{r}
ggplot(systems_ga, aes(x= PWS_TYPE_CODE))+
  geom_bar (fill="steelblue")+
  labs(
    title = "Count of Public Water Sytem Types in Georgia",
    x = "Water System Type",
    y = "Number of Systems"
  )
```


```{r}
systems_ga |>
  count(PWS_TYPE_CODE, sort = TRUE)
  
```

This allows for summary statistics to be calculate, in this context, this shows the distirbution and range of populations served by public water systems in Georgia

```{r}
summary (systems_ga$POPULATION_SERVED_COUNT)
  
```


This puts into persepctive of how many people that the water systems serve. Approximately 25% of the systems served fewer than 49 people, while 75% of the systems served fewer than 262 people. 

```{r}
systems_ga_filtered <- systems_ga |>
  filter(POPULATION_SERVED_COUNT>0)
ggplot(
  data=systems_ga_filtered,
  mapping = aes(x=POPULATION_SERVED_COUNT)
) +
  geom_histogram(bins=40)+
  scale_x_log10(
    breaks = c(10,100,1000,10000,100000),
    labels = c("10", "100", "1,000", "10,000", "100,000")
  ) +
  labs(
    title = "Distribution of Population Served by Public Water Systems in Georgia",
    x= "Population Served",
    y= "Number of Systems"
  )
  
```

In this step, I filtered the to include public water systems in Georgia that serve a population > 0. After this, a histogram was created to visualize the distribution of population sized serve by the public systems. 

The histogram appear to be a strongly skewed right distribution, showing a large varation in system size. Most water systems serve a smaller population of individuals with the median population served being 100 people. However, the long right tail is a result of some water systems serving a large population of individuals reaching up to 1 million indivduals. 


```{r}
violations <- read_csv(here("data", "raw-data", "SDWA_PN_VIOLATION_ASSOC.csv"))
```

```{r}
library(dplyr)

# Filter water systems in Georgia
systems_ga <- systems %>%
  filter(STATE_CODE == "GA")

# Keep violations for those Georgia water systems
violations_ga <- violations %>%
  inner_join(systems_ga %>% select(PWSID), by = "PWSID")

# Count violations per system in Georgia
violation_summary <- violations_ga %>%
  group_by(PWSID) %>%
  summarise(violation_count = n()) %>%
  count(violation_count) %>%
  rename(Number_of_Systems = n, Violation_Count = violation_count)

print(violation_summary)
```

```{r}
ggplot(violation_summary, aes(x = factor(Violation_Count), y = Number_of_Systems)) +
  geom_bar(stat = "identity", fill = "darkred") +
  labs(
    title = "Distribution of Drinking Water Violations per Public Water System in Georgia",
    x = "Number of Violations",
    y = "Number of Water Systems"
  ) +
  theme_minimal()
```

# Inferential Statistical Analysis( Part 2 )
Moving forward to formally evaluate whether rural public water systems in Georgia are more likely to experience drinking water violations compared to urban systems, a multivariable logistic regression model was fit. The outcome of interest was whether a system experienced at least one violation. The primary predictor of interest was rural versus urban classification. The model additionally adjusted for population served and water system type to account for potential confounding.

## Creating System-Level Violation Counts
This step aggregates violation records to the system level and merges them back into the Georgia water system dataset. Systems with no recorded violations are assigned a count of zero.

```{r}
# Count violations per system
violation_counts <- violations_ga %>%
  group_by(PWSID) %>%
  summarise(violation_count = n())

# Merge violation counts back to data
analysis_data <- systems_ga %>%
  left_join(violation_counts, by = "PWSID") %>%
  mutate(
    violation_count = ifelse(is.na(violation_count), 0, violation_count)
  )
```

## Merging Rural-Urban Classification
This step merges the RUCA ZIP code classification dataset to the system-level dataset. Rural classification was defined using RUCA codes, where codes ≥ 4 were classified as rural and codes 1–3 were classified as urban.
```{r}
analysis_data <- analysis_data %>%
  mutate(ZIP_CODE = as.character(ZIP_CODE))

ruca <- ruca %>%
  mutate(ZIPCode = as.character(ZIPCode))
```

```{r}
# Merge RUCA classification (to RUCS ZIPCode)
analysis_data <- analysis_data %>%
  left_join(ruca, by = c ("ZIP_CODE" = "ZIPCode"))
```

```{r}
# Create rural indicator (based on PrimaryRUCA codes, zipcodes without match code are left as NA (n=1417)
analysis_data <- analysis_data %>%
  mutate(
    rural = ifelse(PrimaryRUCA >= 4, 1, 0)
  )
```
```{r}
# Check the first few rows 
head(analysis_data %>% select(ZIP_CODE, PrimaryRUCA, rural))

# Check how many rural rows you have
table(analysis_data$rural, useNA = "ifany")
```

Rural indicator (rural) was created using the PrimaryRUCA codes from the RUCA ZIPCode dataset. For ZIP codes in the water systems data that did not have a corresponding RUCA code, the rural value is NA. These NA values are currently kept in the dataset to indicate missing RUCA information. However, we will decided to remove them to ensure accurate rural-urban classification.

```{r}
# Remove systems with missing rural classification
analysis_data <- analysis_data %>%
  filter(!is.na(rural))

# Check updated counts
table(analysis_data$rural)
nrow(analysis_data)
```

## Creating A Binary Outcome System 
We will create a binary outcome variable to indicate whether a water system had at least one recorded drinking water violation (1 = yes, 0 = no). Among the Georgia public water systems included in the sample, 8.5% experienced at least one recorded drinking water violation, while 91.5% had no recorded violations. This indicates that violations were relatively uncommon overall, though measurable subset of systems were affected.
```{r}
# Create binary violation indicator
analysis_data <- analysis_data %>%
  mutate(
    any_violation = ifelse(violation_count > 0, 1, 0)
  )

# Check distribution
table(analysis_data$any_violation)
prop.table(table(analysis_data$any_violation))
```

## Preparing Modeling Variables
Before fitting the logistics regression model, several variable were prepared and transformed
  * Population served was filtered to exlcude systems reporting zero population. Additionally, it was log transofmed to account of the strong right skew observed in the raw population distribution.
  * Rural status, violations, and water system type were coverted to categorical variables.
  
```{r}
# We need our modeling variables 
analysis_data <- analysis_data %>%
  filter(POPULATION_SERVED_COUNT > 0) %>%          # removing systems that report serving n=0
  mutate(
    log_population = log(POPULATION_SERVED_COUNT), # log transformation, orignal graph skewed right
    rural = as.factor(rural),                      # rural as categorical
    any_violation = as.factor(any_violation),      # violation as categorical
    PWS_TYPE_CODE = as.factor(PWS_TYPE_CODE)       # water system as categorical
  )

# summary statistics of our log transformation
summary(analysis_data$log_population)
```

## Fitting the Regression Model
Next, we fit a multivariable logistic regression to evaluate whether rural systems are more likely to have at least one violation, adjusting for population served and system type. The results are summarized by obtaining odds ratios (OR) and 95% confidence intervals (CI):
  * Rural systems (rural=1) had 1.84 times the odds of experiencing at least one violation (95% CI: 1.46, 2.33)
  * Population served presented an odds ratio of 1.15 per unit increase (95% CI:1.08, 1.22) indicating that larger water systems tend to have slightly higher odds of recording a violation.
  * Community water systems comparison (CWS)
    * Non-transient non-community systems (NTNCWS) had lower odds (OR=0.37) of violations (95% CI: 0.21-0.61)
    * Transient non-community systems (TNCWS)
    similarly had lower odds (OR=0.51) of violations (95% CI:0.39-0.68)

Overall, the model indicates that rural systems are at higher risk for drinking water violations when accounting for system type and the population served. 

```{r}
# Are rural systems more likely to have at least one violation
model1 <- glm(
  any_violation ~ rural + log_population + PWS_TYPE_CODE, 
  data = analysis_data,
  family = binomial 
)

# View model summary: coefficients, standard errors, z-values, and p-values
summary(model1)

# Odds ratios
exp_coef <- exp(cbind(
  OR = coef(model1),                       
  confint(model1)             # 95% CI for coefficients
))

exp_coef

# Summarize the logistic regression results
summary(model1)

```

## Generating a Logistic Regression Model
To visualize the results of the logistic regression model, we create a forest plot displaying the odds ratios for each predictor. This allows for a clear comparison of the direction of associations between rural status, population served, system type, and the likelihood of having at least one drinking water violation.
```{r}
# Load required package
library(ggplot2)
library(dplyr)
library(broom)

# Get tidy regression results with ORs and 95% CI
tidy_model <- tidy(model1, exponentiate = TRUE, conf.int = TRUE)

# Remove intercept for plotting, we don't need this
tidy_model_plot <- tidy_model %>%
  filter(term != "(Intercept)")

# Create forest plot of ORs
ggplot(tidy_model_plot, aes(x = term, y = estimate)) +
  geom_point(size = 3, color = "steelblue") +               # OR points
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),    # 95% CI
                width = 0.2, color = "steelblue") +
  coord_flip() +                                           # flip
  labs(
    title = "Odds Ratios for Logistic Regression Predictors",
    x = "Predictor",
    y = "Odds Ratio (95% CI)"
  ) +
  theme_minimal()
```